---
title: "Distance in Stan: group size and density."
author: "Len Thomas"
date: "2026-04-28"
format:
  html:
    number-sections: true
    code-fold: true
  pdf: 
    number-sections: true
    echo: false
    geometry:
      - bottom=3cm
      - top=3cm
      - left=2cm
      - right=2cm
execute: 
  cache: true
---

# Introduction

```{r}
#| code-summary: "Code: load libraries and define some constants"
#| message: false
#| warning: false

#Load some libraries
#library(inlabru) 
library(dsm)
library(ggplot2)
library(sf)
library(knitr)
library(kableExtra)

library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores() - 1)

library(Distance)
library(truncnorm)

set.seed(3024)
```

Thomas (2026) presented Stan code to fit a half-normal detection function to line transect distance sampling data.  Here we extend that code to include group size as a covariate in the detection model and to estimate density using a model-based approach.

We base test runs on the GoMx spotted dolphin data also used by Thomas (2026).

```{r}
#| code-summary: "Code: Load mexdolphinlot group size distribution"

#Load distdata, segdata, obsdata, preddata
data(mexdolphins)

#distdata
#Convert distances from m to km
distdata$distance <- distdata$distance / 1000
#Create a centred size covariate for potential use
distdata$s_size <- distdata$size - mean(distdata$size)

#segdata
#Add count of detections per segment into segdata
segdata$n <- sapply(segdata$Sample.Label, 
                    function(s) sum(obsdata$Sample.Label == s))
#Convert effort from m to km
segdata$l <- segdata$Effort / 1000
```

@fig-group-size shows the distribution of group sizes in the GoMx spotted dolphin data.  I should also probably re-show the detection distances and a map of spatial locations, but for now I move on to fitting code.

```{r}
#| code-summary: "Code: Plot group size distribution"
#| label: fig-group-size
#| fig-cap: "Histogram of group size for the GoMx spotted dolphin data."

#Make a histogram of group size
ggplot(distdata, aes(x = size)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(x = "Group Size", y = "Frequency") +
  theme_minimal()
```

# Stan model

```{r}
#| code-summary: "Code: Stan run params"
stan_file <- "distance_hn_dens.stan"
n_iter <- 3000
n_chains <- 6
```

The Stan code is in the file "`r stan_file`".  The model estimates the scale parameter of the half-normal detection function, the mean and shape parameters of the group size distribution (assumed negative binomial with no spatial structure), and the density of groups (assumed Poisson with no spatial structure) and individuals in the survey area.

Below is the run output, using `r n_chains` chains with `r n_iter` iterations each.

```{r}
#| code-summary: "Code: Stan setup"
#Specify data, parameters to return and initialization function for sigma

w <- 8
stan_data <- list(n = nrow(distdata), 
              x = distdata$distance, 
              w = w, 
              s = as.integer(distdata$size),
              log_sigma_prior_mean = log(w / 2),
              log_sigma_prior_sd = 0.5,
              n_seg = nrow(segdata),
              seg_count = segdata$n,
              seg_l = segdata$l)
#stan_pars <- c("sigma", "p", "mu_s", "phi_s")
#Initialization function for sigma
create_stan_init_f <- function(w){
  stan_init_f <- function(){
    list(sigma = runif(1, w/100, w * 2),
         mu_s = runif(1, 1, 100),
         phi_s = runif(1, 0.1, 1),
         lambda_s = runif(1, 0, max(mean(segdata$n/(segdata$l * 2 * w)))))
  }
  return(stan_init_f)  
}
```
```{r}
#| code-summary: "Code: Fit sim data with Stan"
#| messages: false
#| output: false

#Compile the model
stan_cds_mod <- stan_model(file = stan_file)
#Run it
stan_cds_fit <- sampling(stan_cds_mod,
                         data = stan_data, 
                         init = create_stan_init_f(w),
                         #pars = stan_pars, 
                         iter = n_iter, thin = 1, chains = n_chains)
#samples <- extract(stan_cds_fit)
```

```{r}
#| code-summary: "Code: Summarize results"
print(stan_cds_fit)
```

# Discussion

Some potential next steps:

- lots of model checking needed; evaluation of priors
- add spatial structure to the group size distribution and density models
- add covariates to density model and to detection function model
- extend detection function model to additional key function (hazard rate) or potentially a mixture model (or half-normals) for more flexibility
- change the current segment-based model for encounters to one on a continuous field (like, say, INLA does), which will then allow group-level covariates on the detection function


