---
title: "Gaussian Processes for Spatial Modelling"
subtitle: "A practical introduction"
author: "MMARINeDNA / eDNAVisualJointModel"
date: today
format:
  revealjs:
    embed-resources: true
    theme: simple
    slide-number: true
    progress: true
    history: false
    chalkboard: false
    fig-align: center
    fig-cap-location: bottom
    code-fold: true
    code-line-numbers: false
    html-math-method: katex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  dpi       = 110
)
suppressPackageStartupMessages({
  library(ggplot2)
  library(MASS)
  library(patchwork)
  library(viridis)
  library(tibble)
  library(dplyr)
  library(tidyr)
})
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

## What is a Gaussian process?

A **GP is a probability distribution over functions**.

::: {.fragment}
Every finite collection of values is jointly multivariate normal:

$$
\bigl(f(\mathbf{x}_1),\, \ldots,\, f(\mathbf{x}_N)\bigr) \;\sim\; \mathcal{MVN}(\mathbf{m},\, \mathbf{K}).
$$
:::

::: {.fragment}
Fully specified by mean $m(\mathbf{x})$ and covariance $k(\mathbf{x}, \mathbf{x}')$. We write $f \sim \mathcal{GP}(m, k)$.
:::

## Sample paths from a 1-D GP prior

```{r}
#| fig-width: 9
#| fig-height: 4

se_kernel <- function(x1, x2, sigma = 1, ell = 1) {
  sigma^2 * exp(-0.5 * outer(x1, x2, "-")^2 / ell^2)
}

x_grid <- seq(0, 10, length.out = 200)
n_samples <- 4

draw_samples <- function(ell, sigma = 1) {
  K <- se_kernel(x_grid, x_grid, sigma = sigma, ell = ell) +
       diag(1e-8, length(x_grid))
  L <- chol(K)
  samples <- t(L) %*% matrix(rnorm(length(x_grid) * n_samples),
                             nrow = length(x_grid))
  as_tibble(samples, .name_repair = ~paste0("draw", seq_along(.))) |>
    mutate(x = x_grid) |>
    pivot_longer(-x, names_to = "draw", values_to = "f")
}

ell_values <- c(0.3, 1, 3)
plot_data <- bind_rows(lapply(ell_values, function(ell) {
  draw_samples(ell) |> mutate(ell_lab = sprintf("italic(l) == %.1f", ell))
})) |> mutate(ell_lab = factor(ell_lab,
                               levels = sprintf("italic(l) == %.1f", ell_values)))

ggplot(plot_data, aes(x = x, y = f, colour = draw)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ ell_lab, labeller = label_parsed) +
  scale_colour_viridis_d(option = "D", end = 0.85, guide = "none") +
  labs(title = expression("Sample functions from " *
                          GP(0, k[SE]) *
                          " at three length-scales"),
       x = expression(x), y = expression(f(x))) +
  theme(strip.text = element_text(size = 13, face = "bold"),
        plot.title = element_text(size = 13))
```

::: {.fragment}
Length-scale $\ell$ controls **wiggliness**: small $\ell$ = wiggly, large $\ell$ = smooth.
:::

## The squared-exponential kernel

$$
k_{\mathrm{SE}}(\mathbf{x}, \mathbf{x}') \;=\; \sigma^2 \exp\!\Bigl(-\tfrac{\|\mathbf{x} - \mathbf{x}'\|^2}{2\ell^2}\Bigr).
$$

Two hyperparameters: $\sigma$ = amplitude, $\ell$ = length-scale.

```{r}
#| fig-width: 9
#| fig-height: 3

x1 <- 0
xs <- seq(-5, 5, length.out = 200)
ell_levels <- c("italic(l) == 0.3", "italic(l) == 1.0", "italic(l) == 3.0")
kdf <- bind_rows(
  tibble(d = xs, k = se_kernel(x1, xs, sigma = 1, ell = 0.3)[1, ],
         ell_lab = ell_levels[1]),
  tibble(d = xs, k = se_kernel(x1, xs, sigma = 1, ell = 1.0)[1, ],
         ell_lab = ell_levels[2]),
  tibble(d = xs, k = se_kernel(x1, xs, sigma = 1, ell = 3.0)[1, ],
         ell_lab = ell_levels[3])
) |> mutate(ell_lab = factor(ell_lab, levels = ell_levels))

ggplot(kdf, aes(x = d, y = k, colour = ell_lab)) +
  geom_line(linewidth = 1.0) +
  scale_colour_viridis_d(option = "D", end = 0.8, name = NULL,
                         labels = function(x) parse(text = x)) +
  labs(x = expression(x - {x*minute}),
       y = expression(k(x, {x*minute}))) +
  theme(legend.position = "top")
```

# Gaussian processes for spatial fields {.center}

## The spatial setup

We have a latent field $f(\mathbf{x})$ at coordinates $\mathbf{x}_i$:

$$
\log \lambda(\mathbf{x}) \;=\; \mu \;+\; f(\mathbf{x}),\qquad f \sim \mathcal{GP}(0, k).
$$

::: {.fragment}
Observations are conditional on the latent field:

$$
y_i \mid f(\mathbf{x}_i) \;\sim\; p\bigl(y_i \mid \mu + f(\mathbf{x}_i)\bigr).
$$
:::

::: {.fragment}
Likelihoods we use: qPCR Ct hurdle, ZI-Beta-Binomial reads, Poisson encounter rates.
:::

## Anisotropic kernels for spatial data

One length-scale per dimension:

$$
k(\mathbf{x}, \mathbf{x}') \;=\; \sigma^2 \exp\!\Biggl(-\frac{1}{2}\sum_{d=1}^{D} \frac{(x_d - x'_d)^2}{\ell_d^2}\Biggr).
$$

```{r}
#| fig-width: 9
#| fig-height: 3.4

draw_2d_field <- function(lx, ly, n_grid = 50) {
  grid <- expand.grid(x = seq(0, 10, length.out = n_grid),
                      y = seq(0, 10, length.out = n_grid))
  d2 <- outer(grid$x, grid$x, "-")^2 / lx^2 +
        outer(grid$y, grid$y, "-")^2 / ly^2
  K <- exp(-0.5 * d2) + diag(1e-8, nrow(d2))
  L <- chol(K)
  f <- as.numeric(t(L) %*% rnorm(nrow(grid)))
  grid |> mutate(f = f)
}

# Use parsable plotmath for facet labels. '*' = juxtaposition, '~~' =
# thin space, ':'/';' are quoted literal punctuation.
panel_levels <- c("Isotropic*\":\"~~italic(l)[x] == 2*\";\"~~italic(l)[y] == 2",
                  "Anisotropic*\":\"~~italic(l)[x] == 0.5*\";\"~~italic(l)[y] == 4")
set.seed(7)
df_iso <- draw_2d_field(2, 2)   |> mutate(panel = panel_levels[1])
set.seed(7)
df_aniso <- draw_2d_field(0.5, 4) |> mutate(panel = panel_levels[2])
df_2d <- bind_rows(df_iso, df_aniso) |>
  mutate(panel = factor(panel, levels = panel_levels))

ggplot(df_2d, aes(x = x, y = y, fill = f)) +
  geom_raster() +
  facet_wrap(~ panel, labeller = label_parsed) +
  scale_fill_viridis_c(option = "viridis", name = expression(f(x, y))) +
  coord_equal(expand = FALSE) +
  labs(x = expression(x), y = expression(y)) +
  theme(strip.text = element_text(size = 11, face = "bold"))
```

# Fitting a Gaussian process {.center}

## Exact GP regression

Closed-form posterior at new points $\mathbf{x}_*$:

$$
\boldsymbol{\mu}_* = \mathbf{K}_{*,X}\,(\mathbf{K}_{X,X} + \sigma_n^2 \mathbf{I})^{-1}\,\mathbf{y}.
$$

::: {.fragment}
**The catch**: requires inverting $N \times N$. Cholesky is $\mathcal{O}(N^3)$ — and HMC does it *every leapfrog step*.
:::

::: {.fragment}
For ecological surveys with $N \sim 10^3{-}10^4$, the exact GP is impractical.
:::

## Approximation strategies

The three approaches we've explored:

::: {.incremental}
- **Splines as basis functions** — replace $f(\mathbf{x})$ with a linear combination of pre-chosen spline bases (thin-plate, B-splines, P-splines). Wood (2003); Eilers & Marx (1996).
- **SPDE / Matérn** — Lindgren–Rue–Lindström (2011); the foundation of R-INLA's spatial framework.
- **Spectral / Hilbert-space GP (HSGP)** — Solin & Särkkä (2020); Riutort-Mayol *et al.* (2023). A basis-function method using Laplacian eigenfunctions, with weights from the kernel's spectral density.
:::

::: {.fragment}
This project uses **HSGP** — see next section.
:::

# Hilbert-space approximate GPs {.center}

## HSGP: the idea

Solin & Särkkä (2020): any **stationary** kernel is approximated by

$$
k(\mathbf{x}, \mathbf{x}') \;\approx\; \sum_{j=1}^{M} S\!\bigl(\sqrt{\lambda_j}\bigr)\,\phi_j(\mathbf{x})\,\phi_j(\mathbf{x}'),
$$

with $\phi_j, \lambda_j$ = Laplacian eigenfunctions / eigenvalues on a bounded domain.

::: {.fragment}
Equivalently, a finite **basis-function expansion**:

$$
f(\mathbf{x}) \;\approx\; \sum_{j=1}^{M} \sqrt{S(\sqrt{\lambda_j})}\;\phi_j(\mathbf{x})\,z_j,\qquad z_j \sim \mathcal{N}(0, 1).
$$
:::

## HSGP: the payoff

::: {.incremental}
- $\mathcal{O}(NM)$ per likelihood — linear in $N$.
- Basis matrix $\boldsymbol{\Phi}_{N \times M}$ built once, reused.
- HMC-friendly: $\sigma$, $\ell$, $z_j$ all live in standard Stan machinery.
- Latent $f$ feeds **any** non-Gaussian likelihood without changing the basis.
:::

## HSGP in 1-D — the recipe

On $[-L, L]$ with Dirichlet boundaries:

$$
\phi_j(x) \;=\; \frac{1}{\sqrt{L}} \sin\!\Bigl(\frac{j\pi(x + L)}{2L}\Bigr),\qquad \sqrt{\lambda_j} \;=\; \frac{j\pi}{2L}.
$$

```{r}
#| fig-width: 9
#| fig-height: 2.8

L <- 1.5
x_grid <- seq(-1, 1, length.out = 300)
phi_1d <- function(L, j, x) (1/sqrt(L)) * sin(j * pi * (x + L) / (2 * L))

phi_df <- bind_rows(lapply(1:6, function(j) {
  tibble(x = x_grid, phi = phi_1d(L, j, x_grid),
         j = factor(j, levels = 1:6))
}))

ggplot(phi_df, aes(x = x, y = phi, colour = j)) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_d(option = "D", end = 0.85,
                         name = expression(j)) +
  labs(x = expression("x (normalised)"),
       y = expression(phi[j](x)))
```

## Multi-D HSGP — tensor products

Basis is a **product** of 1-D eigenfunctions:

$$
\phi_{(j_1, \ldots, j_D)}(\mathbf{x}) \;=\; \prod_{d=1}^{D} \phi^{(d)}_{j_d}(x_d).
$$

::: {.fragment}
For an anisotropic SE kernel, the spectral density factors:

$$
S(\boldsymbol{\omega}) \;=\; \sigma^2 (2\pi)^{D/2} \prod_{d=1}^{D} \ell_d \exp\!\Bigl(-\tfrac{1}{2}\ell_d^2\,\omega_d^2\Bigr).
$$
:::

::: {.fragment}
Implementation: $\mathbf{f} = \boldsymbol{\Phi}\,(\mathbf{w} \odot \mathbf{z})$ where $\mathbf{w}$ are the spectral weights.
:::

## How to choose $L$ and $M$

Riutort-Mayol *et al.* (2023):

::: {.fragment}
- **Boundary**: $c = L / S \geq 1.5$ where $S$ is the data half-range.
:::

::: {.fragment}
- **Basis count per dimension**: $m \geq 1.75\,c\,/\,\rho_\ell$ where $\rho_\ell = \ell / S$.
:::

::: {.fragment}
- **Trade-off**: more basis → better approximation, higher cost.
:::

::: {.fragment}
Practical defaults: $c = 1.5$, $m \in [10, 40]$ per dim.
:::

## Summary

::: {.incremental}
- A GP is a distribution over functions, fully specified by $m$ and $k$.
- Length-scales $\ell_d$ encode spatial smoothness — anisotropic for coastal data.
- Exact GPs are $\mathcal{O}(N^3)$; impractical at our $N$.
- **HSGP** swaps the GP for a truncated basis expansion: $\mathcal{O}(NM)$.
- Two dials per dim: $c \geq 1.5$, $m \geq 1.75c/\rho_\ell$.
:::

## Key references

::: {.smaller}
- Williams & Rasmussen (2006). *Gaussian Processes for Machine Learning.* MIT Press.
- Diggle & Ribeiro (2007). *Model-Based Geostatistics.* Springer.
- Lindgren, Rue & Lindström (2011). *JRSS-B* 73:423–498.
- Solin & Särkkä (2020). *Stat. Comp.* 30:419–446.
- Riutort-Mayol *et al.* (2023). *Stat. Comp.* 33:17.
- Heaton *et al.* (2019). *JABES* 24:398–425.

Companion presenter notes: `presentations/gp_models_presenter_notes.html`.
:::
